# Data import
First, the data and the covariates have to be imported.
```{r Data import}
# clear environment
rm(list = ls())

# Library and environment prep
source("/home/rstudio/analysis/Code/R_projects/EPIC_Analysis/Dexa-repo/Library_and_Environment_Dexa.R")

# read in the for match batch corrected M values
GR_corrected_filtered_full <- readRDS(paste0(dataDirectory, "/rds_dexa/GR_corrected_filtered_full.rds"))
GR_corrected_filtered_light <- readRDS(paste0(dataDirectory, "/rds_dexa/GR_corrected_filtered_light.rds"))

# import DMPs with annotation, RnBeads
DMP_CD14_CaCt_anno_gene <- readRDS(paste0(outputDirectory, "/DMP_CD14_CaCt_anno_gene"))

# importing meta data as covariates
covariates <- read_csv(paste0(dataDirectory, "/covariates_Dexa.csv"))

# import EPIC annotation
epic_annotate <- as.data.frame(readRDS(paste0(general_dataDirectory,"/epic_annotate.rds")))
```

# Linear Regression
## Fitting the model
First, the lineear model has to be fitted. To see differences between Non_Responder and Responder for specific cell types, I subsetted those and checked for individual differences in those cell types. 
```{r Fit-linear-model}
# fit the linear model for CD14
mVals_CD14 <- mVals[,grepl("_CD14$", colnames(GR_corrected_filtered_full))]
head(mVals_CD14[,1:5])
bVals_CD14 <- bVals[,grepl("_CD14$", colnames(GR_corrected_filtered_full))]
head(bVals_CD14[,1:5])
design_CD14 <- design[1:20,]
fit_CD14 <- lmFit(mVals_CD14, design_CD14)
```

## Contrast matrix
A contrast matrix is created for a pairwise comparison between Non_Responder and Responder which is then fitted to the linear model. Summary of the linear model showcasing the number of down-/upregulated CpGs is printed.
```{r contrast-matrix-and-fit}
# create a contrast matrix for specific comparisons
contMatrix <- makeContrasts(Non_Responder - Responder,
                            levels=design)
contMatrix

contMatrix_CD14
fit2_CD14 <- contrasts.fit(fit_CD14, contMatrix)
fit2_CD14 <- eBayes(fit2_CD14)

summary(decideTests(fit2_CD14))
```

# Investigate DMPs
## Visualize differentially methylated CpGs 
To the lists with differentially methylated positions (DMPs), the annotation data frame is joined. Furthermore, distributions of p-values of the different cell types are visualized as volcano plots. 
```{r Differentially-methylated-positions, fig.cap = "DMP visualization"}
DMPs_CD14 <- rownames_to_column(topTable(fit2_CD14, num=Inf), "cgid") %>%
  left_join(epic_annotate, by = "cgid")
head(DMPs_CD14)
write.table(DMPs_CD14, file=paste0(outputDirectory,"/DMPs_CD14_corrected_match.csv"), sep=",", row.names=FALSE)
```

## Adding differences to p-values
For further analysis, beta values are added to the DMP data frame of the CD14 samples. Additionally, the combined rank is generated for better comparisons. The combined rank ranks CpGs based on p-value, mean differences and log2 fold difference and keeps the worst of the three ranks. This can be used to rank DMPs.
```{r adding-b-value-differences}
# get the b Values in a table for Non_Responder/Responder
bVals_Ca <- bVals_CD14[,grepl("Non_Responder", colnames(bVals_CD14))] %>%
  as_tibble(rownames = "cgid") %>%
  mutate(mean.Non_Responder = rowMeans(select(.,-cgid)))

bVals_Ct <- bVals_CD14[,grepl("Responder", colnames(bVals_CD14))] %>%
  as_tibble(rownames = "cgid") %>%
  mutate(mean.Responder = rowMeans(select(.,-cgid)))

# get the difference between the b values for plotting
bVals_diff <- left_join(bVals_Ca, bVals_Ct, by = ("cgid")) %>%
  mutate(mean.diff = mean.Non_Responder - mean.Responder) %>%
  select(cgid, mean.Non_Responder, mean.Responder, mean.diff)

# join with the DMP tables  
DMPs_CD14_diff <- DMPs_CD14 %>%
  left_join(bVals_diff, by = "cgid")
head(DMPs_CD14_diff)

# plotVolcano adjusted p values, CD14 mean.diff png
png(paste0(outputDirectory, "/Dexa_png_hq/Volcano_DMPs_CD14_mean_diff.png"), width = 6, height = 5, res = 120, units = "in", family = "ArialMT")
ggplot(DMPs_CD14_diff, aes(x = mean.diff, y = -log10(adj.P.Val))) + 
  geom_point(pch = 21, fill = "darkgrey", color = "black", size = 2) + 
  geom_hline(yintercept = -log10(0.05), col = "red") + 
  scale_y_continuous(limits = c(-0.01, 2)) + 
  ggtitle(paste0("DMPs CD14 Non_Responder vs. Responder")) +
  xlab("mean.diff") +
  ylab("-log10(p_val_adj)") +
  theme(panel.grid.major = element_line("white"),
        panel.grid.minor = element_line("white"),
        axis.title.y = element_text(size = 8),
        axis.text.y = element_text(size = 8),
        axis.title.x = element_text(size = 8),
        axis.text.x = element_text(size = 8),
        legend.title = element_blank(),
        legend.position = "none",
        plot.title = element_text(face = "bold", size = 10))
dev.off()
```

Seeing a bunch of DMPs without context is often overwhelming and unhelpful. A way to see common trends are gene ontology enrichment analysis like GO and KEGG. The classical way is using only the significant CpGs, but this does not account for the whole gene, as these have more than one CpG. Therefore, I use methylGSA which takes all CpGs and asjusts for the number of probes as a covariate. http://www.bioconductor.org/packages/devel/bioc/vignettes/methylGSA/inst/doc/methylGSA-vignette.html  
```{r GO-and-Kegg-term}
# using methylGSA
# methylGSA implements logistic regression adjusting number of probes as a covariate

# extracting p values
DMPs_CD14_P_Values <- DMPs_CD14$P.Value

# name vector with CpG names
names(DMPs_CD14_P_Values) <- DMPs_CD14$cgid

# all CpGs
GO_terms_CD14_BC <- methylglm(DMPs_CD14_P_Values, array.type = "EPIC")
GO_terms_CD14_BC
write_csv(GO_terms_CD14_BC, paste0(outputDirectory, "/GO_terms/GO_terms_CD14_BC.csv"))

# only promoters 
GO_terms_CD14_BC_promoter <- methylglm(DMPs_CD14_P_Values, group = "promoter2", array.type = "EPIC")
GO_terms_CD14_BC_promoter
write_csv(GO_terms_CD14_BC_promoter, paste0(outputDirectory, "/GO_terms/GO_terms_CD14_BC_promoter.csv"))

# KEGG
KEGG_terms_CD14_BC <- methylglm(DMPs_CD14_P_Values, array.type = "EPIC", GS.type = "KEGG")
write_csv(KEGG_terms_CD14_BC, paste0(outputDirectory, "/GO_terms/KEGG_terms_CD14_BC.csv"))
KEGG_terms_CD14_BC_promoter <- methylglm(DMPs_CD14_P_Values, group = "promoter2", array.type = "EPIC", GS.type = "KEGG")
write_csv(KEGG_terms_CD14_BC_promoter, paste0(outputDirectory, "/GO_terms/KEGG_terms_CD14_BC_promoter.csv"))
```
